### 2020-07-09 paper.tex
finish proof for Jacobian

%\usepackage{aistats2020}
% If your paper is accepted, change the options for the package
% aistats2020 as follows:
%
\usepackage[accepted]{aistats2020}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.

@@

\subsection{Distributions on traces}
Define target density $\pi$ (XXX below it is called $p$, fix this) on the joint space $X \times U$, where $X$ are traces of the model and $X$ are traces of the proposal, in terms of the density $p(x)$ and the family of densities $q_x(u)$ as $\pi(x, u) := p(x) q_u(x)$.

@@

\subsection{Stationarity of the involution with respect to joint density}
Let $(Z, \Sigma, \mu)$ be a measure space of pairs of traces of the model and proposal, where $\mu$ is $\sigma$-finite.
Consider a $\mu$-measurable function $h : Z \to Z$ that is an involution ($h(h(z)) = z$) and let $\mu \circ h^{-1}$ denote the pushforward of $\mu$ under $h$, such that $\mu \circ h^{-1}$ is $\sigma$-finite and such that $\mu \circ h^{-1}$ is absolutely continuous with respect to $\mu$.
% TODO can we simplify the conditions, using the fact that h is an involution?
Then, the Radon-Nikodym derivative exists, and is denoted $(d (\mu \circ h^{-1}) / d \mu)(z)$.

The goal is to prove stationarity with respect to the measure defined by density $p(z)$ with respect to $\mu$.
The move starts with state $z \in Z$, applies the involution $h$, and then accepts (returns $h(z)$) with probability $\alpha(z)$ and otherwise returns $z$.
Let $k_{z}$ denote the measure on new states induced by the involution $h$ and the previous state $z$:
\begin{equation}
k_{z}(A) = [h(z) \in A] \alpha(z) + [z \in A] (1 - \alpha(z))
\end{equation}
The stationarity condition is:
\begin{equation}
\int_Z p(z) k_{z}(A) \mu(dz) = p(A) = \int_Z p(z) [z \in A] \mu(dz) \mbox{ for all } A \in \Sigma
\end{equation}
Substituting in the definition of $k_z(A)$ and simplifying gives:
\begin{equation} \label{eq:stationarity-requirement}
\int_Z p(z) [h(z) \in A] \alpha(z) \mu(dz) = \int_Z p(z) [z \in A] \alpha(z) \mu(dz) \mbox{ for all } A \in \Sigma
\end{equation}
Note that for a $\mu$-measurable function $g$ such that $g \circ h$ is integrable with respect to $\mu$:
\begin{equation} \label{eq:pushforward}
\int_Z g(h(z)) \mu(dz) = \int_Z g(z) (\mu \circ h^{-1})(dz) = \int_Z g(z) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \mu(dz)
\end{equation}
For $g(z) := p(h(z)) [z \in A] \alpha(h(z))$ we have $g(h(z)) = p(z) [h(z) \in A] \alpha(z)$, which is the integrand of the left-hand-side of Equation~(\ref{eq:stationarity-requirement}).
Applying Equation~(\ref{eq:pushforward}) to the left-hand side of Equation~(\ref{eq:stationarity-requirement}) gives:
\begin{equation}
\int_Z p(z) [h(z) \in A] \alpha(z) \mu(dz) = \int_Z g(z) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \mu(dz) = \int_Z p(h(z)) [z \in A] \alpha(h(z)) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \mu(dz)
\end{equation}
Equating this with the right-hand side of Equation~(\ref{eq:stationarity-requirement}) gives the following sufficient condition for stationarity:
\begin{equation}
\int_Z p(h(z)) [z \in A] \alpha(h(z)) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \mu(dz) = \int_Z p(z) [z \in A] \mu(dz) \mbox{ for all } A \in \Sigma
\end{equation}
It therefore suffices to find $\alpha$ such that:
\begin{equation}
p(h(z)) [z \in A] \alpha(h(z)) \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) = p(z) [z \in A] \alpha(z) \mbox{ for all } z \in Z
\end{equation}
The following choice of $\alpha$ suffices:
\begin{equation}
\alpha(z) = \min\left\{ 1, \frac{p(h(z))}{p(z)} \left( \frac{d (\mu \circ h^{-1})}{d \mu}(z) \right) \right\}
\end{equation}

All requirements:
\begin{enumerate}
\item $\mu$ is $\sigma$-finite
\item $h$ is $\mu$-measurable
\item $h$ is an involution
\item the pushforward $\mu \circ h^{-1}$ is $\sigma$-finite
\item the pushforward $\mu \circ h^{-1}$ is absolutely continuous with respect to $\mu$
\item the function $g(z) := p(h(z)) [z \in A] \alpha(h(z))$ is $\mu$-measurable for all $A \in \Sigma$
\item the function $(g \circ h)(z) = p(z) [h(z) \in A] \alpha(z)$ is integrable with respect to $\mu$ for all $A \in \Sigma$
\end{enumerate}

\paragraph{Characterizing the pushforward $\mu \circ h^{-1}$ for involution $h$}
Can we guarantee that it is $\sigma$-finite or absolutely continuous with respect to $\mu$?
Are there simpler sufficient conditions we can give?
$(\mu \circ h^{-1})$ is $\sigma$-finite if $Z$ is the countable union of $(mu \circ h^{-1})$-measurable sets with finite measure.

$\sigma$-finite means

\section{Detailed balance}

Can we employ Theorem 2 of Tierney 1998? How does this relate to how the algorithm is described.
Tierney already employs it for the involution-based move for the joint space of model and proposal traces (see page 4, bullet point 2 'deterministic proposals').
But employing it for the composite move that includes the proposal and involution is more complex.
The elements are:
\begin{enumerate}
\item $x, y \in X$ and $(X, \Sigma_X, \mu_X)$, where $x$ are traces of the model, and $\pi(x)$ is the model (target) density.
\item $u \in U$ and $(U, \Sigma_U, \mu_U)$, where $u$ are traces of the proposal and $q_{x}(u)$ is the proposal density.
\item An involution $h : X \times U \to X \times U$ (properly restricted to the support of $\pi(x) q_{x}(u)$.
\end{enumerate}
We want to provide a detailed balance result on $(X, \Sigma_X, \mu_X)$, proving detailed balance with respect to $\pi(x)$.
Tierney's proposal kernel is defined by a family of measures $Q_{x} : \Sigma_X \to [0, \infty)$.
In our case, this is:
\[
Q_{x}(A) := \int_U [h(x, u)_1 \in A] q_{x}(du) = \int_U [h(x, u)_1 \in A] q_{x}(u) \mu_U(du)
\]
Tierney's transition kernel with acceptance probability $\alpha(x \to y)$ is defined by a family of measures $P_x : \Sigma_X \to [0, \infty)$ (Equation 1 of Tierney):
\[
P_x(A) = \int_A \alpha(x \to y) Q_x(dy) + [x \in A] \int_X (1 - \alpha(x \to y)) Q_x(dy) \mbox{ for } A \in \Sigma_X
\]
Detailed balance (Tierney equation 2) is expressed as an equality between two measures on $X \times X$, denoted $(\pi P)_{\to}$ and $(\pi P)_{\gets}$.
\[
(\pi P)_{\to}(A, B) = (\pi P)_{\gets}(A, B)
\]
where
\[
(\pi P)_{\to}(A, B) := \int_A \left( \int_B P_x(dy) \right) \pi(dx)
\]
and
\[
(\pi P)_{\gets}(A, B) := \int_B \left( \int_A P_x(dy) \right) \pi(dx)
\]
Tierney Proposition 1 relies on a $\sigma$-finite measure $\mu$ on the product space $(X \times X, \Sigma_X \times \Sigma_X$), which is constructed as
the measure induced by first sampling $x$ according to $\pi$ and then sampling $y$ according to the proposal $Q_x$:
\[
\mu(A, B) := \int_A \left( \int_B Q_x(dy) \right) \pi(dx) \;\;\; \mu^T(A, B) := \mu(B, A)
\]
In our case, this is:
\[
\mu(A, B) := \int_A \left( \int_U [h(x, u)_1 \in B] q_{x}(du) \right) \pi(dx) = \int_A \left( \int_U [h(x, u) \in B \times U] q_{x}(du) \right) \pi(dx)
\]
\paragraph{Show that $\mu$ is $\sigma$-finite (a requirement for Tierney Proposition 1)}
$X \times X$ needs to be the countable union of measurable sets with finite measure $\mu(X \times X)$.
It should build on the fact that $\pi$ and $q_x$ are $\sigma$-finite.
%Can we use the fact that the composition of two $\sigma$-finite kernels is itself $\sigma$-finite (where the first kernel is $\pi$ and has a trivial index space)?
%Apparently 
%If so, then it would suffice to show that $Q$ is a $\sigma$-finite kernel, which is true if all $Q_x$ are $\sigma$-finite measures.

\paragraph{Characterize $R$, and $\mu_R$ is the restriction of $\mu$ to $R$, and $\mu_R^T$ is as above}

\paragraph{Characterize the following Radon-Nikodym derivative of $\mu_R$ with respect to $\mu_R^T$}
\[
r(x, y) = \frac{d\mu_R}{d\mu_R^T}(x, y)
\]
We need to show that:
\begin{enumerate}
\item $\alpha$ is $\mu$-almost everywhere zero on $R^c$
\item $\alpha(x, y) r(x, y) = \alpha(y, x)$
\end{enumerate}
Then, according to Tierney Proposition 1, there exists a (unique up to null-sets) symmetric set $R \in \Sigma_X \times \Sigma_X$ such that $\mu$ and $\mu^T$ are mutually absolutely continuous.
%(TODO: constructing $R$ for our kernel is a key step).

\subsection{Special case of involution on discrete + diffeomorphism}

\section{Derivation of pushforward Radon-Nikodym derivative for special case}
TODO: Do some special case where it simplifies to a Jacobian determinant.
Maybe the case where the involution is composed of an involution on discrete part and a diffeomorphism on the continuous part.

\section{Proof of detailed balance for involution}
\citet{tierney1998note} gives a class of MCMC kernels based on involutions that satisfy detailed balance.
Specifically, given an involution $T : Z \to Z$, they denote the pushforward of $\pi$ under $T$ by $\pi'$:
$\pi'(A) := \pi(T^{-1}(A))$.
They let $\nu(A) := \pi(A) + \pi'(A)$.
They let $h(z)$ denote `a density for $\pi$ with respect to $\nu$'.
Then, they set:
\[
\alpha(x \to T(x)) \frac{h(x)}{h(T(x))} = \alpha(T(x) \to x)
\]
Our definition of $\alpha$ satisfies:
\[
\frac{\alpha(z \to T(z))}{\alpha(T(z) \to z)} = \frac{\pi(T(z))}{\pi(z)} \left( \frac{d (\mu \circ T^{-1})}{d \mu} (z) \right)
\]
Therefore, to use Tierney's result, it suffices to show that (when $\pi$ and $\pi'$ are $\sigma$-finite and have densities with respect to the same reference measure $\mu$):
\[
\frac{h(T(z))}{h(z)}= \frac{\pi(T(z))}{\pi(z)} \left( \frac{d (\mu \circ T^{-1})}{d \mu} (z) \right)
\]
(note that $T$ is our $f$).
What is $h(z)$, the density of $\pi$ with respect to $\nu(A) := \pi(A) + \pi'(A)$?
Well, if $\pi$ and $\pi'$ have densities with respect to $\mu$, then so does $\nu$, and it is $\nu(z) = \pi(z) + \pi'(z)$, so the density of $\pi$ with respect to $\nu$ is:
\[
h(z) = \frac{\pi(z)}{\nu(z)} %= \frac{\pi(z)}{\pi(z) + \pi'(z)}
\]
and therefore:
\[
\frac{h(T(z))}{h(z)} = \frac{\pi(T(z))}{\pi(z)} \cdot \frac{\nu(z)}{\nu(T(z))}
\]
It suffices to show that:
\[
\frac{\nu(T(z))}{\nu(z)} = \left( \frac{d (\mu \circ T^{-1})}{d \mu} \right)^{-1}
\]
First, note that $\nu$ and $\nu \circ T^{-1}$ are the same measure:
\begin{align}
\nu(A) &= \pi(A) + \pi'(A) = \pi(A) + \pi(T^{-1}(A))\\
(\nu \circ T^{-1})(A) &= \pi(T^{-1}(A)) + \pi(T^{-1}(T^{-1}(A))) = \pi(T^{-1}(A)) + \pi(A)
\end{align}
Therefore,
\[
\frac{d (\nu \circ T^{-1})}{d \nu}(z) = 1 \mbox{ for all } z
\]
Expanding $d(\nu \circ T^{-1}) / d \nu$ using the chain rule:
\begin{align}
1 &= \frac{d (\nu \circ T^{-1})}{d \nu}(z)\\
&=   \frac{d (\nu \circ T^{-1})}{d (\mu \circ T^{-1})}(z) \cdot
    \frac{d (\mu \circ T^{-1})}{d \mu}(z) \cdot
    \frac{d \mu}{d \nu}(z)\\
&= \nu(T(z))  \frac{d (\mu \circ T^{-1})}{d \mu} \frac{1}{\nu(z)}
\end{align}
where in the second step we used that
\[
\frac{d (\nu \circ T^{-1})}{d (\mu \circ T^{-1})}(z) = \frac{d \nu}{d \mu}(T(z)) = \nu(T(z))
\]
To show this, use (i) $\nu \circ T^{-1} = \nu$, (ii) the definition of Radon-Nikodym derivative $d \nu / d \mu$ and (iii) and Theorem 3.6.1 in Bogachev with $g := d\nu / d\mu$:
\begin{align}
(\nu \circ T^{-1})(A) = \nu(A) &= \int_A \frac{d \nu}{d \mu}(z) \mu(dz) = \int_A \frac{d \nu}{d \mu} (T(z)) ( \mu \circ T^{-1} )(dz)
\end{aling}

\section{Proof of stationarity for involution}
Stationarity follows from detailed balance by setting $B := Z$ in the detailed balance equation, and using $k'_z(Z) = \alpha(z) + (1 - \alpha(z)) =1$, so the right-hand side simplifies to $\int_A \pi(dz) = \pi(A)$.

\section{Proof of stationarity for end-to-end kernel}
We are given that the involution is stationary with respect to (the measure induced by) $\pi(x, u) := p(x) q_x(u)$:
\[
\int_Z k'_z(A) \pi(z) \mu(dz) = \pi(A) \mbox{ for all } A \in \Sigma_Z
\]
It suffices to show that the end-to-end kernel $k$ is stationary with respect to (the measure induced by) $p(x)$:
\[
\int_X k_x(A) p(x) \mu_X(dx) = p(A) \mbox{ for all } A \in \Sigma_X
\]
The kernel $k$ on $X$ defined by Algorithm~\ref{alg:involutive-mcmc} for valid $p$, $q$, and $f$:
\[
k_x(A) := \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \;\; \mbox{for all} \;\; A \in \Sigma_X, x \in X
\]
is stationary with respect to $p$.
\begin{align*}
    \int_X k_x(A) p(x) \mu_X(dx) &= p(A) \mbox{ for all } A \in \Sigma_X\\
    \int_X \left( \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
    \int_{X \times U} k'_{x,u}(A \times U) q_x(u) p(x) \mu(dz) &= p(A)\\
    \int_{X \times U} k'_{x,u}(A \times U) \pi(z) \mu(dz) &= p(A)\\
    \pi(A \times U) &= p(A)\\
    \int_{A \times U} q_x(u) p(x) \mu(dz) &= p(A)\\
    \int_A \left( \int_U q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
    \int_A p(x) \mu_X(dx) &= p(A)
\end{align*}
% TODO characterize the various assumptions that allow us to e.g. rearrange the order of integration, etc.

>>>>

\usepackage{aistats2020}
% If your paper is accepted, change the options for the package
% aistats2020 as follows:
%
%\usepackage[accepted]{aistats2020}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.

@@

\subsection{Distributions on traces}
% TODO rewrite this section..
Define target density $\pi$ (XXX below it is called $p$, fix this) on the joint space $X \times U$, where $X$ are traces of the model and $X$ are traces of the proposal, in terms of the density $p(x)$ and the family of densities $q_x(u)$ as $\pi(x, u) := p(x) q_u(x)$.

@@

\subsection{Derivation of the pushforward Radon-Nikodym derivative for a special case}
We consider the special case in which the involution $f$ can be factored into an involution on a countable set $I$ and a family of bijections on $\mathbb{R}^{n_i}$ for some $n_i$ for each $i \in I$.
Suppose $Z = \{(i, \mathbf{x}) : i \in I, \mathbf{x} \in \mathbb{R}^{n_i}\}$.
Suppose $f_1$ is an involution on $I$ and $n_i = n_{f_1(i)}$ and $f_2$ is a family of continuously differentiable bijections indexed by $i \in I$, such that $f_{2,i} : \mathbb{R}^{n_i} \to \mathbb{R}^{n_i}$.
Also suppose that $f_{2,i} = f_{2,f_1(i)}^{-1}$ for all $i \in I$. That is,
\begin{equation}
f_{2,i}(f_{2,f_1(i)}(\mathbf{x})) = \mathbf{x} \mbox{ for all } \mathbf{x} \in \mathbb{R}^{n_i} \mbox{ and all } i \in I
\end{equation} 
Then, $f : Z \to Z : (i, \mathbf{x}) \mapsto (f_1(i), f_{2,i}(\mathbf{x}))$ is an involution because:
\begin{equation}
f(f(i, \mathbf{x})) = f(f_1(i), f_{2,i}(\mathbf{x})) = (f_1(f_1(i)), f_{2,f_1(i)}(f_{2,i}(\mathbf{x}))) = (i, \mathbf{x})
\end{equation}
Let $\Sigma_n$ and $\mu_n$ denote the Lebesgue $\sigma$-algebra and Lebesgue measure on $\mathbb{R}^n$, respectively.
Let $\Sigma \subset \mathcal{P}(Z)$ be the $\sigma$-algebra of sets of the form $\cup_{i \in I} \{(i, \mathbf{x}) : \mathbf{x} \in A_i\}$ for some $A_i \in \Sigma_{n_i}$ for each $i \in I$.
%Let $\Sigma$ be the $\sigma$-algebra on $Z$ generated by the sets $G := \{ \{(i, \mathbf{x}) : \mathbf{x} \in A\} : i \in I, A \in \Sigma_{n_i} \}$.
%Note that the sets $G$ form a $\pi$-system:
%\begin{align}
%\{(i, \mathbf{x}) : \mathbf{x} \in A\} \cap \{(j, \mathbf{x}) : \mathbf{x} \in B\} = \varnothing = \{(i, \mathbf{x}) : \mathbf{x} \in \varnothing \} \in G \mbox{ for } i \ne j\\
%\{(i, \mathbf{x}) : \mathbf{x} \in A\} \cap \{(i, \mathbf{x}) : \mathbf{x} \in B\} = \{(i, \mathbf{x}) : \mathbf{x} \in A \cap B\} \in G \mbox{ because } A \cap B \in \Sigma_{n_i}
%\end{align}
Let $\mu$ denote the measure on measurable space $(Z, \Sigma)$ given by:%, uniquely defined by its value on the sets $G$:
\begin{equation}
\mu(\cup_{i \in I} \{(i, \mathbf{x}) : \mathbf{x} \in A_i\}) := \sum_{i \in I} \mu_{n_i}(A_i)
\end{equation}
We wish to show that the Radon-Nikodym derivative of the pushforward of $\mu$ by $f$ with respect to $\mu$, evaluated at $(i, \mathbf{x})$, is the absolute value of the Jacobian (determinant) of the function $f_{2,i}$ evaluated at $\mathbf{x}$, which is denoted $(J f_{2,i})(\mathbf{x})$:
\begin{equation}
\frac{d (\mu \circ f^{-1})}{d \mu}(i, \mathbf{x})
= \left| (Jf_{2,i})(\mathbf{x})\right|
\end{equation}
Consider $(\mu \circ f^{-1})(A)$ for $A \in \Sigma$:
\begin{align}
(\mu \circ f^{-1})(\cup_{i \in I} \{ (i, \mathbf{x}) : \mathbf{x} \in A_i \})
&= \mu(f^{-1}(\cup_{i \in I} \{ (i, \mathbf{x}) : \mathbf{x} \in A_i \}))\\
&= \mu(\cup_{i \in I} f^{-1}(\{ (i, \mathbf{x}) : \mathbf{x} \in A_i \}))\\
&= \mu(\cup_{i \in I} \{ (f_1(i), \mathbf{x}) : \mathbf{x} \in f_{2,i}(A_i) \})\\
&= \sum_{i \in I} \mu_{n_i}(f_{2,i}(A_i))
\end{align}
It suffices to show that for all $A \in \Sigma$:
\begin{equation}
\int_A \left| (Jf_{2,i})(\mathbf{x})\right| \mu(dz) = \sum_{i \in I} \mu_{n_i}(f_{2,i}(A_i))
\end{equation}
Expanding the left-hand side:
\begin{align}
\int_A \left| (Jf_{2,i})(\mathbf{x})\right| \mu(dz)
= \sum_{i \in I} \int_{A_i} \left| (Jf_{2,i})(\mathbf{x})\right| \mu_{n_i}(d \mathbf{x}) % TODO how to justify this? it seems obvious
= \sum_{i \in I} \mu_{n_i}(f_{2,i}(A_i)) % Borgachev Theorem 4.7.1 
\end{align}
where the final step uses Borgachev Theorem 3.7.1 with $g := 1$, and $F := f_{2,i}$.

\subsection{Proof of detailed balance for involution}
\citet{tierney1998note} gives a class of MCMC kernels based on involutions that satisfy detailed balance.
Specifically, given an involution $T : Z \to Z$, they denote the pushforward of $\pi$ under $T$ by $\pi'$:
$\pi'(A) := \pi(T^{-1}(A))$.
They let $\nu(A) := \pi(A) + \pi'(A)$.
They let $h(z)$ denote `a density for $\pi$ with respect to $\nu$'.
Then, they set:
\[
\alpha(x \to T(x)) \frac{h(x)}{h(T(x))} = \alpha(T(x) \to x)
\]
Our definition of $\alpha$ satisfies:
\[
\frac{\alpha(z \to T(z))}{\alpha(T(z) \to z)} = \frac{\pi(T(z))}{\pi(z)} \left( \frac{d (\mu \circ T^{-1})}{d \mu} (z) \right)
\]
Therefore, to use Tierney's result, it suffices to show that (when $\pi$ and $\pi'$ are $\sigma$-finite and have densities with respect to the same reference measure $\mu$):
\[
\frac{h(T(z))}{h(z)}= \frac{\pi(T(z))}{\pi(z)} \left( \frac{d (\mu \circ T^{-1})}{d \mu} (z) \right)
\]
(note that $T$ is our $f$).
What is $h(z)$, the density of $\pi$ with respect to $\nu(A) := \pi(A) + \pi'(A)$?
Well, if $\pi$ and $\pi'$ have densities with respect to $\mu$, then so does $\nu$, and it is $\nu(z) = \pi(z) + \pi'(z)$, so the density of $\pi$ with respect to $\nu$ is:
\[
h(z) = \frac{\pi(z)}{\nu(z)} %= \frac{\pi(z)}{\pi(z) + \pi'(z)}
\]
and therefore:
\[
\frac{h(T(z))}{h(z)} = \frac{\pi(T(z))}{\pi(z)} \cdot \frac{\nu(z)}{\nu(T(z))}
\]
It suffices to show that:
\[
\frac{\nu(T(z))}{\nu(z)} = \left( \frac{d (\mu \circ T^{-1})}{d \mu} \right)^{-1}
\]
First, note that $\nu$ and $\nu \circ T^{-1}$ are the same measure:
\begin{align}
\nu(A) &= \pi(A) + \pi'(A) = \pi(A) + \pi(T^{-1}(A))\\
(\nu \circ T^{-1})(A) &= \pi(T^{-1}(A)) + \pi(T^{-1}(T^{-1}(A))) = \pi(T^{-1}(A)) + \pi(A)
\end{align}
Therefore,
\[
\frac{d (\nu \circ T^{-1})}{d \nu}(z) = 1 \mbox{ for all } z
\]
Expanding $d(\nu \circ T^{-1}) / d \nu$ using the chain rule:
\begin{align}
1 &= \frac{d (\nu \circ T^{-1})}{d \nu}(z)\\
&=   \frac{d (\nu \circ T^{-1})}{d (\mu \circ T^{-1})}(z) \cdot
    \frac{d (\mu \circ T^{-1})}{d \mu}(z) \cdot
    \frac{d \mu}{d \nu}(z)\\
&= \nu(T(z))  \frac{d (\mu \circ T^{-1})}{d \mu} \frac{1}{\nu(z)}
\end{align}
where in the second step we used that
\begin{equation}
\frac{d (\nu \circ T^{-1})}{d (\mu \circ T^{-1})}(z) = \frac{d \nu}{d \mu}(T(z)) = \nu(T(z))
\end{equation}
To show this, we use the definition of the Radon-Nikodym derivative $d (\nu \circ T^{-1}) / d (\mu \circ T^{-1})$.
To show that $z \mapsto (d \nu / d \mu)(T(z))$ is this Radon-Nikodym derivative, it suffices to show that for all measurable $A$:
\begin{equation}
\int_A \left( \frac{d \nu}{d \mu} (T(z)) \right) ( \mu \circ T^{-1} )(dz) = (\nu \circ T^{-1})(A) := \nu(T^{-1}(A))
\end{equation}
Applying Theorem 3.6.1 in Bogachev with $Y := A$, $y := z$, $x := z'$, $X := T^{-1}(A)$, $f := T$, and $g(y) := (d\nu / d\mu)(T(y))$:
\begin{align}
\int_A \left( \frac{d \nu}{d \mu} (T(z)) \right) ( \mu \circ T^{-1} )(dz)
&= \int_Y g(y) ( \mu \circ f^{-1} )(dy)\\
&= \int_X g(f(x)) \mu(dx) \;\;\;\; [\mbox{Bogachev Theorem 3.6.1}]\\
&= \int_{T^{-1}(A)} (d\nu / d\mu)(T(T(z'))) \mu(dz')\\
&= \int_{T^{-1}(A)} (d\nu / d\mu)(z') \mu(dz')\\
&= \int_{T^{-1}(A)} \nu(dz')
= \nu(T^{-1}(A))
\end{align}
% NOTE Bogachev's theorem uses domains of integration which are the whole
% sample spaces for the two sigma algebras.  I think that the theorem should
% also apply to any measurable sets A and T(A) because one can construct a
% sigma algebra and measure that are restricted to to any nonempty measurable
% set.

\subsection{Proof of stationarity for involution}
Stationarity follows from detailed balance by setting $B := Z$ in the detailed balance equation, and using $k'_z(Z) = \alpha(z) + (1 - \alpha(z)) =1$, so the right-hand side simplifies to $\int_A \pi(dz) = \pi(A)$.

\subsection{Proof of stationarity for end-to-end kernel}
We are given that the involution is stationary with respect to (the measure induced by) $\pi(x, u) := p(x) q_x(u)$:
\[
\int_Z k'_z(A) \pi(z) \mu(dz) = \pi(A) \mbox{ for all } A \in \Sigma_Z
\]
It suffices to show that the end-to-end kernel $k$ is stationary with respect to (the measure induced by) $p(x)$:
\[
\int_X k_x(A) p(x) \mu_X(dx) = p(A) \mbox{ for all } A \in \Sigma_X
\]
The kernel $k$ on $X$ defined by Algorithm~\ref{alg:involutive-mcmc} for valid $p$, $q$, and $f$:
\[
k_x(A) := \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \;\; \mbox{for all} \;\; A \in \Sigma_X, x \in X
\]
is stationary with respect to $p$.
\begin{align*}
    \int_X k_x(A) p(x) \mu_X(dx) &= p(A) \mbox{ for all } A \in \Sigma_X\\
    \int_X \left( \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
    \int_{X \times U} k'_{x,u}(A \times U) q_x(u) p(x) \mu(dz) &= p(A)\\
    \int_{X \times U} k'_{x,u}(A \times U) \pi(z) \mu(dz) &= p(A)\\
    \pi(A \times U) &= p(A)\\
    \int_{A \times U} q_x(u) p(x) \mu(dz) &= p(A)\\
    \int_A \left( \int_U q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
    \int_A p(x) \mu_X(dx) &= p(A)
\end{align*}
% TODO characterize the various assumptions that allow us to e.g. rearrange the order of integration, etc.
